The Stakes of Scrying: \\ \lstinline[style=inlinecode]{.^} dotket and the Seer Proposal

### 2024-04-25 mss.tex
Final cut.

The Nock 12 opcode and the \dotket~rune are used by user-\newline{}space code to ask the kernel for data in the scry namespace.\footnote{The scry namespace, in brief, is a collection of immutable values “bound” by convention at fixed endpoints.  The scry namespace is defined by a unique path for each resource.  A representative scry path would have the entries \texttt{/host/rift/life/vane/request-type/revision-number/desk\newline{}/file-path/mark}.}  When Urbit runs userspace code, that code is not raw Nock.  It is instead called Mock, which is Nock with two additions:   deterministic errors are caught and returned as values to the kernel; and there is an extra Nock opcode, opcode 12.  A simple Nock 12 formula looks like \lstinline[style=inlinecode]{[12 [1 /foo/bar/baz]]}, where \lstinline[style=inlinecode]{/foo/bar/baz} is a path literal \lstinline[style=inlinecode]{[%foo %bar %baz ~]}, representing a path in the scry namespace.  The result of running this opcode will be the result of performing the scry request, i.e. dereferencing the path.

Catching deterministic errors is uncontroversial, but the Nock 12 opcode poses some difficulties and has been under scrutiny for years now.  The most tangible problem with Nock 12 is that it prevents userspace code from persistently memoizing Nock computations; at the moment, only kernelspace code can hold onto Nock memoization caches across multiple Arvo events.  The lifetime of a userspace memoization cache is limited to a single metacircular evaluation context, i.e. a single call to \lstinline[style=inlinecode]{++mink} or an equivalent gate that runs Mock (virtual Nock with the extra opcode).

@@

\noindent
While error handling is not relevant to the current question, a brief description is included here for thoroughness.  The \lstinline[style=inlinecode]{error} in \lstinline[style=inlinecode]{++mink}'s result \lstinline[style=inlinecode]{[2 error]} is a deterministic error, meaning that every attempt to evaluate that \lstinline[style=inlinecode]{[subject formula]} pair will always result in that same error value—deterministic errors correspond to the lines in the Nock spec where \lstinline[style=inlinecode]{*a} evaluates to \lstinline[style=inlinecode]{*a}, i.e. a trivial infinite loop.  Nondeterministic errors, such as running out of usable memory or getting interrupted by the user, result in the function failing to complete, and the Arvo event will be aborted and rolled back to its state before trying to run this new event.

@@

A scry block means the system refused to answer the question, i.e. the request was "blocked" for some reason—this could be due to lack of permission, or because the request asked about the future, or the request asked for data from another ship that isn't synchronously available within this Arvo event. If the scry blocks, \lstinline[style=inlinecode]{++mink} returns a \lstinline[style=inlinecode]{[1 block]}, where \lstinline[style=inlinecode]{block} is the scry request path whose value could not be resolved.

@@

  \item  Have the interpreter record the set of scry \lstinline[style=inlinecode]{[path value]} pairs looked up by the computation.  This would allow userspace code to perform scry requests safely, but the cost is a nontrivial amount of machinery in the runtime, along with a computational cost of looking up a cache line that grows linearly with the number of scry paths that were requested.  Of the three options listed here, this is the most reasonable, but it is still highly nontrivial and the requirement to compare all requested paths could be onerous.
\end{enumerate}

\noindent
Instead of implementing any of these options, Urbit has currently simple disabled the ability for userspace code to modify the Nock memoization cache.  This is enforced by the jet for \lstinline[style=inlinecode]{++mink}; the jet is simply a call back into the normal Nock interpreter, which is really a Mock interpreter, whose state includes a list of nouns representing the stack of scry handler gates for the current execution context.  If this list is null \lstinline[style=inlinecode]{~}, then we must be in the kernel and the code is allowed to modify the memoization cache; otherwise, we are in some kind of Mock context, and the code is not allowed to modify the cache.

@@

\noindent
(In this example, use is made of \lstinline[style=inlinecode]{++rapt}.  \lstinline[style=inlinecode]{++rapt} is a monadic scry binding gate \citep{Levan2023b}.)

@@

The current reference implementation of Seer is available at \texttt{urbit/urbit} \citetpr{Levan2023}.  A more complete example is available at \citealt{Levan2023a}.

@@

A Gall agent will be able to emit a new kind of move, in addition to all the moves it can emit now, that asks Gall to perform a scry request on the agent's behalf.  When Gall processes this move, it will perform the scry request and convert the result of the scry into a response move that will be delivered back into the agent in a later activation of the Gall vane, in accordance with the normal move processing order that Gall and Arvo use.  Thus an agent could emit a list of moves such as \lstinline[style=inlinecode]{[poke-1 scry-A poke-2 scry-B poke-3 ~]} and Gall will handle all of those moves in the same order as if they were all pokes, i.e. the kernel will not introduce any exceptions to its normal move processing order for scry request moves.

@@

There is a general consensus among Urbit core developers that a new move-based scry interface is better suited as the lowest layer of the scry system than the Seer monadic interface.  The fact that Seer or \dotket\footnote{Or at least an opt-in form of Mock, rather than it being the global default as it is now.}~could be implemented on top of moves (but not necessarily the other way around) militates in favor of moves at the lowest layer.  Another benefit of this arrangement is that by making all scrys asynchronous, the same interface can be used for both local and remote scrys (since remote scrys are always asynchronous).  Such an abstraction over local vs. remote scry calls removes what would otherwise be an unavoidable branch point in userspace I/O code.  It also matches the rest of Gall's interprocess communication model, which does not distinguish between local and remote.  A strong argument can be made that this is also the most ``grug-brained'' option, which is not to be undervalued.

There remain concerns about application development patterns if \dotket~is removed.  Losing synchronous read access to other parts of the system potentially reduces the ergonomic wins of a naturally growing transaction across applications.  This can be implemented in a way that is almost synchronous for local scrys, and in particular it is still possible to perform multiple scry requests at the same system snapshot, meaning the requesting agent will see a consistent view of the system without any tearing.

Gall will also need to run the \lstinline[style=inlinecode]{++on-peek} arm of a Gall agent in the Seer monad, i.e. that arm will return either \lstinline[style=inlinecode]{[%done scry-result]} to indicate it's done or \lstinline[style=inlinecode]{[%scry scry-request-path callback]} to ask Gall to perform a scry request, in which case Gall performs the scry, injects the result into the callback gate, and repeats, until a \lstinline[style=inlinecode]{%done} is returned.  This should not be a direct problem, but betokens complexity creeping into the Gall interface.

>>>>

The Nock 12 opcode and the \dotket~rune are used by user-\newline{}space code to ask the kernel for data in the scry namespace.\footnote{The scry namespace, in brief, is a collection of immutable values “bound” by convention at fixed endpoints.  The scry namespace is defined by a unique path for each resource.  A representative scry path would have the entries \texttt{/host/rift/life/vane/request-type/revision-number/desk\newline{}/file-path/mark}.}  When Urbit runs userspace code, that code is not raw Nock.  It is instead called Mock, which is Nock with two additions:   deterministic errors are caught and returned as values to the kernel; and there is an extra Nock opcode, opcode 12.  A simple Nock 12 formula looks like \lstinline[style=inlinecode]{[12 [1 /foo/bar/baz]]}, where \lstinline[style=inlinecode]{/foo/bar/baz} is a path literal \lstinline[style=inlinecode]{[%foo %bar %baz ~]}, representing a path in the scry namespace.  The result of running this opcode will be the result of performing the scry request, i.\,e.\ dereferencing the path.

Catching deterministic errors is uncontroversial, but the Nock 12 opcode poses some difficulties and has been under scrutiny for years now.  The most tangible problem with Nock 12 is that it prevents userspace code from persistently memoizing Nock computations; at the moment, only kernelspace code can hold onto Nock memoization caches across multiple Arvo events.  The lifetime of a userspace memoization cache is limited to a single metacircular evaluation context, i.\,e.\ a single call to \lstinline[style=inlinecode]{++mink} or an equivalent gate that runs Mock (virtual Nock with the extra opcode).

@@

\noindent
While error handling is not relevant to the current question, a brief description is included here for thoroughness.  The \lstinline[style=inlinecode]{error} in \lstinline[style=inlinecode]{++mink}'s result \lstinline[style=inlinecode]{[2 error]} is a deterministic error, meaning that every attempt to evaluate that \lstinline[style=inlinecode]{[subject formula]} pair will always result in that same error value—deterministic errors correspond to the lines in the Nock spec where \lstinline[style=inlinecode]{*a} evaluates to \lstinline[style=inlinecode]{*a}, i.\,e.\ a trivial infinite loop.  Nondeterministic errors, such as running out of usable memory or getting interrupted by the user, result in the function failing to complete, and the Arvo event will be aborted and rolled back to its state before trying to run this new event.

@@

A scry block means the system refused to answer the question, i.\,e.\ the request was "blocked" for some reason—this could be due to lack of permission, or because the request asked about the future, or the request asked for data from another ship that isn't synchronously available within this Arvo event. If the scry blocks, \lstinline[style=inlinecode]{++mink} returns a \lstinline[style=inlinecode]{[1 block]}, where \lstinline[style=inlinecode]{block} is the scry request path whose value could not be resolved.

@@

  \item  Have the interpreter record the set of scry \lstinline[style=inlinecode]{[path value]} pairs looked up by the computation.  This would allow userspace code to perform scry requests safely, but the cost is a nontrivial amount of machinery in the runtime, along with a computational cost of looking up a cache line that grows linearly with the number of scry paths that were requested.  Of the three options listed here, this is the most reasonable, but it is still highly nontrivial and the requirement to compare all requested paths could be onerous.
\end{enumerate}
\noindent
Instead of implementing any of these options, Urbit has currently simple disabled the ability for userspace code to modify the Nock memoization cache.  This is enforced by the jet for \lstinline[style=inlinecode]{++mink}; the jet is simply a call back into the normal Nock interpreter, which is really a Mock interpreter, whose state includes a list of nouns representing the stack of scry handler gates for the current execution context.  If this list is null \lstinline[style=inlinecode]{~}, then we must be in the kernel and the code is allowed to modify the memoization cache; otherwise, we are in some kind of Mock context, and the code is not allowed to modify the cache.

@@

\noindent
(In this example, use is made of \lstinline[style=inlinecode]{++rapt}.  \lstinline[style=inlinecode]{++rapt} is a monadic scry binding gate (\citetpr{Levan2023b}, 2023c).)

@@

The current reference implementation of Seer is available at \texttt{urbit/urbit} \citetpr{Levan2023} (2023a).  A more complete example is available at \citetpr{Levan2023a} (2023b).

@@

A Gall agent will be able to emit a new kind of move, in addition to all the moves it can emit now, that asks Gall to perform a scry request on the agent's behalf.  When Gall processes this move, it will perform the scry request and convert the result of the scry into a response move that will be delivered back into the agent in a later activation of the Gall vane, in accordance with the normal move processing order that Gall and Arvo use.  Thus an agent could emit a list of moves such as \lstinline[style=inlinecode]{[poke-1 scry-A poke-2 scry-B poke-3 ~]} and Gall will handle all of those moves in the same order as if they were all pokes, i.\,e.\ the kernel will not introduce any exceptions to its normal move processing order for scry request moves.

@@

There is a general consensus among Urbit core developers that a new move-based scry interface is better suited as the lowest layer of the scry system than the Seer monadic interface.  The fact that Seer or \dotket\footnote{Or at least an opt-in form of Mock, rather than it being the global default as it is now.}~could be implemented on top of moves (but not necessarily the other way around) militates in favor of moves at the lowest layer.  Another benefit of this arrangement is that by making all scries asynchronous, the same interface can be used for both local and remote scries (since remote scries are always asynchronous).  Such an abstraction over local vs. remote scry calls removes what would otherwise be an unavoidable branch point in userspace I/O code.  It also matches the rest of Gall's interprocess communication model, which does not distinguish between local and remote.  A strong argument can be made that this is also the most ``grug-brained'' option, which is not to be undervalued.

There remain concerns about application development patterns if \dotket~is removed.  Losing synchronous read access to other parts of the system potentially reduces the ergonomic wins of a naturally growing transaction across applications.  This can be implemented in a way that is almost synchronous for local scries, and in particular it is still possible to perform multiple scry requests at the same system snapshot, meaning the requesting agent will see a consistent view of the system without any tearing.

Gall will also need to run the \lstinline[style=inlinecode]{++on-peek} arm of a Gall agent in the Seer monad, i.\,e.\  that arm will return either \lstinline[style=inlinecode]{[%done scry-result]} to indicate it's done or \lstinline[style=inlinecode]{[%scry scry-request-path callback]} to ask Gall to perform a scry request, in which case Gall performs the scry, injects the result into the callback gate, and repeats, until a \lstinline[style=inlinecode]{%done} is returned.  This should not be a direct problem, but betokens complexity creeping into the Gall interface.
